### 2016-07-06 motivation.tex
Fig. 4 redone.

In this section, we motivate our ideas via an example written in
\txnimp - a C-like imperative language equipped with a \C{txn} lexical
block defining a transaction scope. Each \C{txn} block is associated
with an indentifier in angle braces that uniquely identifies the
transaction. We use Hoare triple notation to annotate programs with
pre- and post- conditions.

@@

Consider an implementation of a banking application that admits
concurrent withdraw transactions on an account balance (\C{B}), as
shown in Fig.~\ref{fig:motiv-eg-1}. If the initial balance (\C{k}) in
the account is enough to perform both withdraws, then the final
balance, after both transactions commit, is expected to reflect the
effects of both withdraws. The pre- and post-conditions in
Fig.~\ref{fig:motiv-eg-1} reflect this expectation. Indeed, invariants
are guaranteed to hold if both withdraw transactions are serialized,
making \iso{Serializable} isolation ({\sc ser}) level a sufficent
condition to preserve invaraints. But, is {\sc ser} necessary?

@@

Thinking in terms of anomalies, as described above, is how database
programmers are often encouraged to reason about weak isolation.
Unfortunately, such reasoning does not rest on any sound foundation,
and is thus highly error-prone. Reasoning in terms of how weak
isolation variants are implementated is no better since it requires
programmers to understand low-level implementation details of the
database that are far removed from the application semantics. An
attractive  alternative in this context would be a principled
reasoning approach that combines declarative reasoning about isolation
guarantees with operational reasoning about programs. We demonstrate
how our proof system makes this possible in the context of the current
example.

@@

\begin{figure}
\centering
  \includegraphics[scale=0.4]{Figures/motiv-eg-1-hb}

\caption{\small $\hbZ$ arrows allowed by {\sc si} (solid) and {\sc rc} (solid
  and dashed) isolation levels for the example in
  Fig.~\ref{fig:motiv-eg-1}. Each arrow represents an $\hbZ$ relationship
  among all operations of \C{Wd2} and operations in \C{Wd1} that follow
  the arrow head. Letters R, W and C stand for read, write and commit,
  respectively.}
\label{fig:motiv-eg-1-hb}
\vspace*{-8pt}
\end{figure}

@@

Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ from \C{Wd2}
to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All arrows are
legal under $\psi_{RC}$ because in no case does an operation $\eta_1$
from \C{Wd1} happen-before $\eta_2$ of \C{Wd2} without the commit of
\C{Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
all dashed arrows, because dashed arrows establish $\hbZ$ edges
between actions in \C{Wd2} and only a subset of the operations in
\C{Wd1}.  Since the result of an operation (\eg a read) depends on
what happens before that operation, the structure of the $\hbZ$
relation ultimately dictates the final result of the program. The
solid arrow and the first (from left) dashed arrow in
Fig.~\ref{fig:motiv-eg-1-hb} denote $\hbZ$ relationships that do not
affect the value of \C{B} in a way that causes the program to violate
its post-condition. The last two dashed arrows denote $\hbZ$
relationships that lead to the violation of the post-condition. The
task of the reasoning framework is to determine if all $\hbZ$
relationships allowed by an isolation level lead to the satisfaction
of the post-condition.

>>>>

In this section, we motivate our ideas via an example written in
\txnimp - a C-like imperative language equipped with a \C{txn} lexical
block defining a transaction scope. Each \C{txn} block is associated
with an identifier in angle braces that uniquely identifies the
transaction. We use Hoare triple notation to annotate programs with
pre- and post- conditions.

@@

Consider an implementation of a banking application that admits
concurrent withdraw transactions on an account balance (\C{B}), as
shown in Fig.~\ref{fig:motiv-eg-1}. If the initial balance (\C{k}) in
the account is enough to perform both withdraws, then the final
balance, after both transactions commit, is expected to reflect the
effects of both withdraws. The pre- and post-conditions in
Fig.~\ref{fig:motiv-eg-1} reflect this expectation. Indeed, invariants
are guaranteed to hold if both withdraw transactions are serialized,
making \iso{Serializable} isolation ({\sc ser}) level a sufficient
condition to preserve invariants. But, is {\sc ser} necessary?

@@

Thinking in terms of anomalies, as described above, is how database
programmers are often encouraged to reason about weak isolation.
Unfortunately, such reasoning does not rest on any sound foundation,
and is thus highly error-prone. Reasoning in terms of how weak
isolation variants are implemented is no better since it requires
programmers to understand low-level implementation details of the
database that are far removed from the application semantics. An
attractive  alternative in this context would be a principled
reasoning approach that combines declarative reasoning about isolation
guarantees with operational reasoning about programs. We demonstrate
how our proof system makes this possible in the context of the current
example.

@@

\begin{figure}
\centering
\subcaptionbox {
  \label{fig:hb-eg-1}
} [
  0.25\columnwidth
] {
  \includegraphics[scale=0.6]{Figures/hb-eg-1}
}
\subcaptionbox {
  \label{fig:hb-eg-2}
} [
  0.25\columnwidth
] {
  \includegraphics[scale=0.6]{Figures/hb-eg-2}
}
\subcaptionbox {
  \label{fig:hb-eg-3}
} [
  0.25\columnwidth
] {
  \includegraphics[scale=0.6]{Figures/hb-eg-3}
}
\subcaptionbox {
  \label{fig:hb-eg-4}
}{
  \includegraphics[scale=0.6]{Figures/hb-eg-4}
}

\caption{\small $\hbZ$ arrows allowed by $\psi_{SI}$
 and $\psi_{RC}$ specifications
% (Figs.~\ref{fig:hb-eg-1},~\ref{fig:hb-eg-2},~\ref{fig:hb-eg-3},~\ref{fig:hb-eg-4})
for the example in Fig.~\ref{fig:motiv-eg-1}. Each
arrow represents an $\hbZ$ relationship between all operations of
\C{Wd2} and an operation in \C{Wd1}. Letters R, W and C stand for
read, write and commit, respectively.}
\label{fig:motiv-eg-1-hb}
\vspace*{-8pt}
\end{figure}

@@

Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ from \C{Wd2}
to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. Arrows in all
executions are legal under $\psi_{RC}$ because in no execution does an
operation $\eta_1$ from \C{Wd1} happen-before $\eta_2$ of \C{Wd2}
without the commit of \C{Wd1} also happening before $\eta_2$.
$\psi_{SI}$ however disallows $\hbZ$ arrows (hence executions) in
Figs.~\ref{fig:hb-eg-2},~\ref{fig:hb-eg-3},~\ref{fig:hb-eg-4} because
such arrows establish $\hbZ$ edges
between (operations in) \C{Wd2} and only a subset of the operations in
\C{Wd1}. Since the result of an operation (\eg a read) depends on
what happens before that operation, the structure of the $\hbZ$
relation ultimately dictates the final result of the program. The
arrows in Figs.~\ref{fig:hb-eg-1} and~\ref{fig:hb-eg-2} denote
$\hbZ$ relationships that do not affect the value of \C{B} in a way
that causes the program to violate its post-condition. In contrast,
arrows in Figs.~\ref{hb-eg-3} and~\ref{hb-eg-4} denote $\hbZ$
relationships that lead to the violation of the post-condition. The
task of the reasoning framework is to determine if all $\hbZ$
relationships allowed by an isolation level lead to the satisfaction
of the post-condition.
